Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-02-17 analysis/paper/paper.qmd
Edited by temporal pattern results & tried CV

After the first discovery at the Seokjangri site at the 1960s, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. We selected stemmed points that were found during field surveys as well as those were from sites that were never dated but were associated with other Late Palaeolithic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

We distinguished between multiple assemblages at a single site where numerous excavations have been conducted in different locations at the site, and by different institutions under the same site name. For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions. There are six different excavation locations that range from a few meters apart to a few kilometers. Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits. As a result, in this research we used four assemblages from Suyanggae. We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location. Hajinri is the sixth excavation location at Suyanggae, which is 3.5 kilometers apart from the other areas and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka). According to the excavation reports for Hajinri, stemmed points first appeared there around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018]. While we include data from Hajinri here, we consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time. We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points. Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].

@@

To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials and artifact size. To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points. We separated the research time period into three phases based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during and after. We explored the relationship between TCSA values and the location of assemblages by comparing the distributions of TCSA values across ecological and vegetation zones.

@@

Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic, people mostly used quartzite and vein quartz for stone artifacts. Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values of `r n_raw` artifacts with raw material information to examine the interaction of raw material types and TCSA values. We categorized rare raw materials, which have less than 10 artifacts, as "Other".

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to distinguish different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be measured rapidly and objectively [@fenenga1953weights; @shea2006origins]. We then explored the relationship between weight and raw materials on TCSA values. Excluding points that we were not able to directly measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.

@@

We used radiocarbon ages to explore the temporal patterns of the likely uses of stemmed points. After excluding assemblages that have no radiocarbon dates, we used `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time. We divided the artifacts from dated assemblages into three groups based on the LGM: before, during and after, to examine the impact of this major climate event on TCSA values.

@@

We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values. Among `r n_sites` sites, we combined those that contain fewer than five stemmed points and named them as "Other". We explored possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones. We compared distributions across four vegetation zones, and also across 16 zones of different geographical characteristics.

For the vegetation zones we used @yi2010vegetation's divisions of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT). These divisions are based on @yim1975distribution's forest vegetation map, defined by recent temperature and precipitation values.

We also explored spatial patterns using 16 geographical zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship]. The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone. We recognize that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.

@@

The process of the introduction of new technologies can vary depending on the environmental and social contexts of the transmission of the manufacturing techniques for the new tools. Moreover, the new tools themselves can alter the typical contexts of tool use. For example, the introduction of the bow and arrow in American enabled individuals to work without hunting in teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing]. Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups carefully selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip]. Thus, the selection of weapon-tip types is likely to reflect the socio-environmental circumstances that people encountered and managed. For example, @eren2022plains used TCSA to compare the morphological variance of Clovis and Folsom points and claimed that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape. Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing. Folsom points, on the other hand, show a narrower range of variation, indicating a they were more likely used for specific tasks.

@@

We predict a temporal pattern of more variable TCSA values at the first appearance of stemmed points, suggesting that people were using the tools as part of an adaptation to moving into unfamiliar landscapes, with unpredictable variation in patch productivity and travel times. We predict a further increase in the variability of TCSA values in the LGM as lower temperatures alter the distribution of resources and lower the predictability of resource encounters. After the LGM we predict a reduction in the range of TCSA values, with higher temperatures and increased bioproductivity.

We predict that TCSA values vary across vegetation types and geographical zones in response to heterogeneously distributed food resources. In harsh environments with lower patch productivity, similar to the case of LGM duration, we predict a further increase in the variability of TCSA values. On the other had, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey. In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary. Our prediction is that the range of TCSA values will be narrower in coastal areas such as vegetation zones ST and SW and geographical zone ESC, compared to inland areas such as vegetation zone CT and geographical zone CI.

The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi.org/10.17605/osf.io/dqna8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational]. All of the figures, tables, statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

@fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely uses. Overall we see a wide variation of TCSA values. The sample mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`. According to the TCSA ranges presented in @tbl-tcsa-ranges, Korean stemmed points are overall mostly in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. Among the weapon-tip types, only poison arrow tips appear to be absent from these Korean assemblages.

@@

@fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed. Acidic volcanic rocks tend to have skewed lower TCSA values. The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale. The category of "Other" raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks. Overall, there is no clear pattern of TCSA values among different raw materials.

@@

@@

@fig-tcsa-radio-carbon shows `r n_sites_with_radiocarbon_ages` assemblages with stemmed points in chronological order (panel A). Assemblages discarded before LGM tend to possess more stemmed points, such as Suyanggae. These assemblages have a wide range of TCSA values, indicating multiple likely usages of stemmed points. Overall, there is no clear pattern in these assemblages over time. The TCSA range varies depending on the assemblages.

Panel B of @fig-tcsa-radio-carbon shows the distribution of TSCA values from assemblages dating to before the LGM, during the LGM, and after, to explore the impact of climate change on the probably function of the artifacts. Our results show that stemmed points were made predominantly before LGM and only a few left after LGM. Each category shows wide variation in LGM, which indicates diverse uses of stemmed points. While the median TSCA value for assemblages discarded during the LGM is higher than assemblages from earlier and later, there is no statistically significant difference in TCSA values across the three periods (`r three_time_p_pretty`).

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses of stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point use? We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e. width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations. Drawing on evolutionary theory, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed. We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.

>>>>

After the first discovery at the Seokjangri site in the 1960s, more than 450 stemmed points have been found at over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. We included stemmed points discovered during field surveys as well as those found at sites that were never dated but were associated with other Late Palaeolithic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

We distinguished between multiple assemblages at a single site where numerous excavations have taken place in different locations at the site, and by different institutions under the same site name. For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions. There are six different excavation locations that range from a few meters apart to a few kilometers. Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits. As a result, in this research we used four assemblages from Suyanggae. We separated one assemblage from the four by using a different site name, Hajinri, following the convention established by the excavators at that location. Hajinri is the sixth excavation location at Suyanggae, which is 3.5 kilometers apart from the other areas and dated much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka). According to the excavation reports for Hajinri, stemmed points first appeared there around 42 ka with the earliest blades and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018]. While we include data from Hajinri here, we consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time. We have more confidence in finds from Bonggok, which has the second-earliest dates (around 41.5 ka) in our collection, as the first appearance of stemmed points. Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].

@@

To answer our research questions about the likely uses for stemmed points, we calculated TCSA values for the stemmed points in our sample. We also explored the interaction of TCSA values with raw materials and artifact size, which we used weight as a proxy. Because the shape of stone artifacts is highly influenced by the raw materials used [@mcpherron2014experimental]. It is also assumed that raw materials are highly correlated with tool size. We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values. We separated the research time period into three phases based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during, and after the LGM. We explored the relationship between TCSA values and the location of assemblages by comparing the distributions of TCSA values across ecological and vegetation zones.

@@

Selective use of raw materials is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic, people mostly used quartzite and vein quartz for stone artifacts. Then, finer grained materials were added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values of `r n_raw` artifacts with raw material information to examine the interaction of raw material types and TCSA values. We categorized rare raw materials, which have less than 10 artifacts, as "Other".

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy for identifying different types of armatures [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for the overall size of the stemmed points. As a reliable discriminator between tools of different sizes as well as a descriptive attribute, weight can be measured rapidly and objectively [@fenenga1953weights; @shea2006origins]. We then explored the relationship between weight and raw materials on TCSA values. Excluding points that we were not able to directly measure or obtain records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.

@@

We used radiocarbon ages to investigate the temporal patterns of the likely uses of stemmed points. After excluding assemblages that have no radiocarbon dates, we used `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time. We divided the artifacts from dated assemblages into three groups based on the LGM: before, during and after, to examine the impact of this major climate event on TCSA values.

@@

We summarized the distribution of TCSA values for each assemblage to analyze spatial patterns in TCSA values. Among `r n_sites` sites, we combined those that contain fewer than five stemmed points and named them "Other". We explored the possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones. We compared the distributions across four vegetation zones and 16 zones that differed in terms of their geographical characteristics.

The vegetation zones are based on @yi2010vegetation's classification of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT). These divisions are based on @yim1975distribution's forest vegetation map, defined by recent temperature and precipitation values.

We also explored spatial patterns using 16 geographical zones based on geographical characteristics including inland, coastal areas, major rivers, islands, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship]. The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). Then, we used longitudes and latitudes to locate individual assemblages and examined TCSA values for each zone. We fully acknowledge that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.

@@

The process of introducing novel technologies can vary based on the social and environmental context in which the transmission of manufacturing techniques for the new tools takes place. Furthermore, the new tools themselves can alter the typical contexts of tool use. In America, for example, the advent of bows and arrows provided non-elite hunters with the opportunity to produce their own subsistence or pursue individual wealth without the necessity of hunting in teams [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing]. Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups carefully selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip]. Thus, the selection of weapon-tip types is likely to reflect the socio-environmental circumstances that people encountered and managed. In one example, @eren2022plains compared the morphological variance of Clovis and Folsom points and claimed that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscapes. Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing. Folsom points, on the other hand, show a narrower range of variation, indicating they were more likely used for specific tasks.

@@

We predict a temporal pattern of more variable TCSA values at the first appearance of stemmed points, suggesting that the tools were being used as part of an adaptation to moving into unfamiliar landscapes, with unpredictable variations in patch productivity and travel times. The variability of TCSA values in the LGM is predicted to increase as lower temperatures alter the distribution of resources and reduce the predictability of resource encounters. After the LGM we anticipate a reduction in the range of TCSA values, with higher temperatures and increased bioproductivity.

In response to heterogeneously distributed food resources, TCSA values are predicted to vary across vegetation types and geographical zones. In harsh environments with lower patch productivity, similar to the case of LGM duration, we expect to observe a further increase in the variability of TCSA values. On the other hand, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey. In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary. Therefore, our prediction is that the range of TCSA values will be narrower in coastal areas such as vegetation zones ST and SW and geographical zone ESC, compared to inland areas such as vegetation zone CT and geographical zone CI.

The entire R code [@core2021team] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/10.17605/osf.io/dqna8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational]. All of the figures, tables, statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.

@@

@fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely uses. Overall we see a wide variation in TCSA values. The sample mean of TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and the standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`. According to the TCSA ranges presented in @tbl-tcsa-ranges, Korean stemmed points are mostly found in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. Among the weapon-tip types, only poison arrow tips appear to be absent from these Korean assemblages.

@@

@fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed. The TCSA values of acidic volcanic rocks are commonly skewed lower. Other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale. The category of "Other" raw materials includes porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks. Overall, there is no clear pattern of TCSA values among different raw materials.

@@

############ test CV ########### to present
library(raster)
# cv for LGM for example
data_lgm <- tcsa_site_ass_mis %>% 
  filter(mis_stage=='LGM')

cv_lgm <- cv(data_lgm$min_tcsa)

@@

@fig-tcsa-radio-carbon shows `r n_sites_with_radiocarbon_ages` assemblages with stemmed points in chronological order (panel A). Assemblages discarded before LGM tend to possess more stemmed points, such as Suyanggae. These assemblages have a wide range of TCSA values, indicating multiple likely uses for stemmed points. Overall, there is no clear pattern in these assemblages over time. The TCSA range varies depending on the assemblage.

In panel B, @fig-tcsa-radio-carbon shows the distribution of TSCA values from assemblages dating to before, during, and after the LGM, in order to explore the impact of climate change on the likely uses of the artifacts. Our results show that stemmed points were made predominantly before LGM and only a few left after LGM. Each category shows wide variation in LGM, which indicates diverse uses of stemmed points. While the median TSCA value for assemblages discarded during the LGM is higher than assemblages from earlier and later, there is no statistically significant difference in TCSA values across the three periods (`r three_time_p_pretty`).

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses for stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point use? We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e. width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations. Drawing on evolutionary theory, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed. We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.
